---
title: "Pitfalls of density plots"
author: "Teemu Säilynoja"
date: "2023-02-22"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
format:
  html:
    page-layout: full
    toc: true
    code-fold: true
    code-tools: true
    code-line-numbers: true
---

```{r}
#| label: imports
#| message: false
library(ggplot2)
library(ggdist)
library(bayesplot)
library(khroma)
library(latex2exp)
library(sfsmisc)
source("../../../code/helpers.R")
source("../../../code/kde_tests.R")

set.seed(37645624)

good_theme <- bayesplot::theme_default(base_family = "Sans", base_size = 14) + theme(
  axis.text = element_text(colour = "#666666", size = 12),
  axis.ticks = element_line(colour = "#666666"),
  title = element_text(colour = "#666666", size = 16),
  plot.subtitle = element_text(colour = "#666666", size = 14),
  legend.text = element_text(colour = "#666666", size = 12),
  legend.title = element_text(colour = "#666666", size = 14),
  axis.line = element_line(colour = "#666666"))

theme_set(good_theme)
```


```{r}
ecdf_difference_limits <- ecdf_confidence_intervals(
  gamma = adjust_gamma_optimize(1000,1000, .95),
  N = 1000, K = 1000)

x0 <- 0:1000 / 1000

p0 = ggplot(mapping = aes(x = x0)) +
  geom_step(aes(y = (ecdf_difference_limits$lower - 0:1000) / 1000)) +
  geom_step(aes(y = (ecdf_difference_limits$upper - 0:1000) / 1000)) +
  labs(title = "ECDF Difference",
       subtitle = "95% outer simultaneous confidense intervals",
       x = "PIT",
       y = "ECDF - Difference")

```
## Intro

Below, we examine two common pitfalls of density plots made using kernel density estimates (KDEs).
Overall, KDEs have satisfactory performance in summarising observations of continuous random variables,
but, as highlighted below, when the variable is bounded, or contains point masses, the density estimate
can misrepresent the data without any issues being immediately apparent from the density plot.

## Simple continuous example

Sample $x_1,\dots,x_N$, from Laplace distribution with mean, $\mu = 0$, and scale, $b=1$, that is,
the density function of $x$ is $f(x) = \frac{1}{2}\exp\left(-|x|\right)$.
```{r}
#| label: sample 1
x1 <- sample(c(-1,1), 1000, replace = T) * rexp(1000)
```

```{r}
p1 <- ggplot() +
  stat_density(aes(x = x1), bw = "SJ", geom = "line") +
  xlab(TeX("$x$")) +
  ylab("KDE") +
  ggtitle("Laplace distribution",subtitle = "N = 1000\nx ~ 1/2 exp(-|x|)") +
  scale_y_continuous(expand = c(0,NA))

p1
```

The ECDF difference plot doesn't suggest any issues with the KDE
```{r}
#| message: false
p0 + geom_step(aes(y = ecdf(pit_from_densityplot(p1, 1, x1))(x0) - x0))
```


## Bounded continuous data

Sample as above, but the distribution is truncated to the interval $[-1.5,0.5]$.

```{r}
#| label: sample 2
x2 <- x1
while (!all(x2 > -1.5, x2 < .5)) {
  mask <- (x2 < -1.5 | x2 > .5)
  x2[mask] = sample(c(-1,1), sum(mask), replace = T) * rexp(sum(mask))
}
```

We see a difference in the tails of the densityplots, when the boundedness is taken into account in the KDE computations.
```{r}
p2 <- ggplot() + 
  stat_density(aes(x = x2, colour = "Bounded"), geom = "line", bounds = c(-1.5, .5)) +
  stat_density(aes(x = x2, colour = "Naive"), geom = "line") + 
  labs(title = "Bounded observation", subtitle = "KDE computed with and without the bounds",
       x = TeX("$x$"), y = "KDE", colour = "") +
  scale_colour_vibrant() +
  scale_y_continuous(expand = c(0,NA))

p2
```

The ECDF difference plot also suggests calibration issues with the unbounded KDE. 
```{r}
#| label: ecdf 2
#| message: false
p0 + geom_step(aes(y = ecdf(pit_from_densityplot(p2, 2, x2))(x0) - x0,
                   colour = "Naive")) +
  geom_step(aes(y = ecdf(pit_from_densityplot(p2, 1, x2))(x0) - x0, colour = "Bounded")) +
  labs(colour = "") +
  scale_colour_vibrant()
```


## Continuous data with point masses

Simulate an example, where measurements under a certain threshold are reported as zeros.
Here $x\sim\exp(x)$ with all values sampled below $0.1$ reported as zero. 
```{r}
#| label: sample 3
x3 <- rexp(1000, 1)
x3[x3 < .1] <- 0
```

If the measurement threshold is known, the data can be visualized as a mixture of
a discrete point mass and a continuous distribution. The "naive" KDE seems to 
under estimate the density near $0.1$, and can't show the spike at zero. 

Below, the density of the continuous mixture component is scaled by $P(x \neq 0)$
to make the area of the bar and the KDE together add to one.
```{r}
p3 <- ggplot() + 
  geom_rect(
    aes(xmin = 0, xmax = .1, ymin = 0, ymax = 10 * sum(x3 == 0) / length(x3), 
        colour = "Mixture", fill = "Mixture"),
    alpha = .3) +
  stat_density(
    aes(x = x3[x3 > 0], y = (1 - sum(x3 == 0) / length(x3)) * after_stat(density), 
        colour = "Mixture", fill = "Naive"), alpha = 0, bounds = c(.1, Inf), outline.type = "upper") +
  stat_density(aes(x = x3, colour = "Naive", fill = "Naive"), alpha = 0, linewidth = 1, outline.type = "upper") +
  labs(title = "Observation with point mass",
       subtitle = "KDE for full observation and only continuous part",
       x = TeX("x"), y = "", colour = "", fill = "") +
  scale_colour_vibrant() + scale_fill_vibrant() +
  scale_y_continuous(expand = c(0,NA))

p3
```

The calibration of the KDE for the continuous part should be inspected separately,
as the scaled density above would show miscalibration due to missing part of the
observation distribution.  
```{r}
p4 <- ggplot() + 
  stat_density(
    aes(x = x3[x3 > 0], 
        colour = "Mixture"), bounds = c(.1, Inf),
    geom = "line")
```

Again, the ECDF difference plot suggest issues with the naive approach and thus 
could be used to warn user if the naive density plot is drawn.
```{r}
#| label: ecdf 3
#| message: false
p0 + geom_step(aes(y = ecdf(pit_from_densityplot(p3, 3, x3))(x0) - x0, colour = "Naive")) +
  geom_step(aes(y = ecdf(pit_from_densityplot(p4, 1, x3[x3 > 0]))(x0) - x0, 
                colour = "Mixture")) +
  labs(colour = "") +
  scale_colour_vibrant()
```




