---
title: "PPC: Ordinal predictions"
subtitle: "Bayesian network model for predicting the number of pregnancies in an IVF treatment."
date: "2023-02-01"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
author: "Teemu Säilynoja"
image: "../../docs/case_studies/ordinal/bn_classifier_files/figure-html/ppc_bars-1.svg"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    code-line-numbers: true
    default-image-extension: svg
    fig-format: svg

---


```{r imports_and_options, message = FALSE}
library(cmdstanr)
library(DirichletReg)
library(bayesplot)
library(ggplot2)
library(reliabilitydiag)
library(khroma)

# Source for the modified reliability plot
source("../../code/helpers.R")

good_theme <- bayesplot::theme_default(base_family = "Sans") + theme(
  axis.text = element_text(colour = "#666666", size = 12),
  axis.ticks = element_line(colour = "#666666"),
  title = element_text(colour = "#666666", size = 16),
  plot.subtitle = element_text(colour = "#666666", size = 14),
  legend.text = element_text(colour = "#666666", size = 12),
  legend.title = element_text(colour = "#666666", size = 14),
  axis.line = element_line(colour = "#666666"))

theme_set(good_theme)
bayesplot_theme_set(good_theme)
color_scheme_set(scheme = c(unname(colour("vibrant")(7)[c(3,2,5,4,1,6)])))

scale_colour_discrete = scale_colour_vibrant
scale_fill_discrete = scale_fill_vibrant

SEED <- 236543
set.seed(SEED)
SAVE_FITS = TRUE
SIM_IIRM = TRUE # whether to use the inferred parameter values from the paper.

```

# Data generation

```{r data}
standata <- list()
standata$P = 0
while(max(standata$P) < 3) {
  pA = c(rdirichlet(1, rep(3, 3)))
  if (SIM_IIRM == TRUE) {
    pU = c(.78, .58, .26) # IIRM value from paper
  } else {
    pU = sort(runif(3), decreasing = T)
  }
  pS = c(rdirichlet(1, rep(3 , 4)))
  if (SIM_IIRM == TRUE) {
    pE = c(0, .07, .21, .39)# IIRM value from paper
  } else {
    pE = c(0, sort(runif(3)))
  }
  N <- 388
  standata$N = N
  standata$A = sample(1:3, size = N, replace = T, prob = pA)
  standata$S = t(replicate(n = N, sample(1:4, size = 3, replace = T, prob = pS)))
  standata$P = sapply(1:N, function(n) rbinom(1,1, pU[standata$A[n]]) * sum(rbinom(3,1, pE[standata$S[n, ]])))
}
```

# Bayesian network model
Stan implementation of the BN model introduced in the paper. In short, the number of pregnancies is modeled to be equal to the number of viable embryos, $e$, if the uterus is receptive, $u$.
The observed variables are the patient age, $A$ split to 3 age categories, and the quality of the transferred embryo, $S \in \{1,\dots,4\}$, where $1$ means no transfer and categories $2$ to $4$ are in a ascending order based on quality.

$$
P = \mathbb I(U = u)\sum_{i=1}^3\mathbb I(E_i = e),
$$
```{r load_model}
model <- cmdstan_model(stan_file = "../../code/stan-models/bn_classifier.stan")
model
```

The model samples very quickly with no apparent issues during the sampling process.

```{r fit_model}
fit <- model$sample(data = standata,
                    parallel_chains = 4,
                    refresh = 0,
                    seed = SEED)
```

```{r ppm_computation}
# Compute the posterior means of the predictive probability mass of each case
ppm <- matrix(colMeans(fit$draws(variables = "ppm", format = "matrix")), ncol = 4)
```

# Posterior predictive checks

A crude visualization of the posterior frequency of each case with 90% posterior predictive intervals. 
```{r}
#| label: ppc_bars
#| dev.args : 
#|   bg: transparent
ppc_bars(y = standata$P,
         yrep = fit$draws(variables = "yrep", format = "matrix") - 1,
         prob = .9,
         freq = F)
```


```{r}
#| label: reldiag_1
#| dev.args : 
#|   bg: transparent
plot_dotted_reliabilitydiag(x = pmin(1, rowSums(ppm[,-1])),
                y = as.numeric(standata$P != 0), quantiles = 50) +
  labs(title = "Calibration: P >= 1")
```


```{r}
#| label: reldiag_2
#| dev.args : 
#|   bg: transparent
plot_dotted_reliabilitydiag(x = pmin(1, rowSums(ppm[,-c(1,2)])),
                y = as.numeric(standata$P > 1), quantiles = 50) +
  labs(title = "Calibration: P >= 2")
```

```{r}
#| label: reldiag_3
#| dev.args : 
#|   bg: transparent
plot_dotted_reliabilitydiag(x = pmin(1, ppm[,c(4)]), y = as.numeric(standata$P == 3), quantiles = 20) +
  labs(title = "Calibration: P = 3")
```


# Marginal posteriors

## Parameters of interest
```{r}
#| label: mcmc_areas_pU
#| dev.args : 
#|   bg: transparent
color_scheme_set(scheme = color("BuRd")(13)[6:1])
mcmc_areas(fit$draws(variables = "pU", format = "matrix")) + vline_at(v = pU, colour = "#666666")
```

```{r}
#| label: mcmc_areas_pE
#| dev.args : 
#|   bg: transparent
mcmc_areas(fit$draws(variables = "pE", format = "matrix")[,-1]) + vline_at(v = pE[-1], colour = "#666666")
```


## Population parameters

For some reason, the authors also model the frequencies of the different
observation categories in the population. These are recovered quite well.

```{r}
#| label: mcmc_areas_pA
#| dev.args : 
#|   bg: transparent
mcmc_areas(fit$draws(variables = "pA", format = "matrix")) + vline_at(v = pA, colour = "#666666")
```


```{r}
#| label: mcmc_areas_pS
#| dev.args : 
#|   bg: transparent
mcmc_areas(fit$draws(variables = "pS", format = "matrix")) + vline_at(v = pS, colour = "#666666")
```



