---
title: "Calibrarion of Ordinal Posterior Predictions"
author: "Teemu Säilynoja"
date: "2023-02-04"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
image: "../../docs/case_studies/ordinal/ordinal_files/figure-html/reliability-diagrams_1-1.svg"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    code-line-numbers: true
    default-image-extension: svg
    fig-format: svg
# knitr:
#   opts_chunk:
#     classes : "dark-light"

---

```{r}
#| label: options
#| code-summary: "Imports & options"
#| output: false
#| message: false

library("bayesplot")
library("cmdstanr")
library("ggplot2")
library("khroma")
library("quartoExtra")


# Source for the modified reliability plot
source("../../code/helpers.R")

good_theme <- bayesplot::theme_default(base_family = "Sans") + theme(
  axis.text = element_text(colour = "#666666", size = 12),
  axis.ticks = element_line(colour = "#666666"),
  title = element_text(colour = "#666666", size = 16),
  plot.subtitle = element_text(colour = "#666666", size = 14),
  legend.text = element_text(colour = "#666666", size = 12),
  legend.title = element_text(colour = "#666666", size = 14),
  axis.line = element_line(colour = "#666666"))

theme_set(good_theme)
bayesplot_theme_set(good_theme)
color_scheme_set(scheme = c(unname(colour("vibrant")(7)[c(3,2,5,4,1,6)])))

scale_colour_discrete = scale_colour_vibrant
scale_fill_discrete = scale_fill_vibrant


# darkmode_theme_set(
#     dark = ggthemes::theme_stata(scheme = "s1rcolor"),
#     light = ggthemes::theme_stata(scheme = "s1color")
# )


SEED <- 236543
set.seed(SEED)
SAVE_FITS = TRUE
```

This notebook highlights posterior predictive visualizations when the 
posterior predictive distribution is ordinal.

As shown below, the ordinal nature of the predictions allows us to use the
cumulative posterior predictive mass function to assess the calibration of the 
posterior.

## Data set
Below, I use synthetic data generated from observations of $K$ Gaussians with
varying means.

$$\begin{align}
x_n &\sim \mathcal N\!\!\left(k, 0.5^2\right), &\text{for } n \in\{1,\dots,N\}\\
k &\sim \text{Categorical}(\theta_k),&\\
\theta_k &= \frac 1 K, &\text{for } k \in \{1, \dots, K\}.
\end{align}$$

```{r}
#| label: data_generation
#| code-summary: Data generation
K <- 5
N <- 1500
sigma <- .5
c <- sample(1:K, N, replace = T)
x <- rnorm(N, c, sigma)
standata_gmm <- list(K = K,
                     N = N,
                     x = x,
                     y = c,
                     sigma = sigma)

```

```{r}
#| label: data_densities
#| dev.args : 
#|   bg: transparent
ggplot(data.frame(standata_gmm)) +
  geom_density(aes(x = x,
                   colour = as.factor(y),
                   fill = as.factor(y),
                   group = as.factor(y)), alpha = .5) +
  legend_none() +
  xlab("") + ylab("") + ggtitle(paste("The data", sep = ""))
```

## Model

The data is fit with two models, both structured to first normalize the data and
then fit a Gaussian Mixture Model with K = `r K`` mixture components with known variances.
In the first model, when the data is normalized, I omit to scale the variance of the mixture components accordingly.

```{r}
#| label: model_code
#| code-summary: Read model code

gmm <- cmdstan_model("../../code/stan-models/gmm_classifier.stan")
gmm
```

```{r}
#| warning: false
#| label: fit_models
#| code-summary: run inference using Stan
fit_1 <- tryCatch(
  expr = {readRDS(paste("../../code/stan-models/fits/gmm_classifier_1_",SEED,".RDS", sep=""))},
  error = function(e) {
    fit <- gmm$sample(data = c(standata_gmm, list(correct_sigma = 0)),
                      parallel_chains = 4,
                      refresh = 0,
                      seed = SEED,
                      show_messages = F)
    if (SAVE_FITS) {fit$save_object(
      paste("../../code/stan-models/fits/gmm_classifier_1_",SEED,".RDS", sep=""))}
    return(fit)},
  finally = {message("Finished model 1.")})

fit_2 <- tryCatch(
  expr = readRDS(paste("../../code/stan-models/fits/gmm_classifier_2_",SEED,".RDS", sep="")),
  error = function(e) {
    fit <- gmm$sample(data = c(standata_gmm, list(correct_sigma = 1)),
               parallel_chains = 4,
               refresh = 0,
               seed = SEED,
               show_messages = F)
    if (SAVE_FITS) {fit$save_object(
      paste("../../code/stan-models/fits/gmm_classifier_2_",SEED,".RDS", sep=""))}
    return(fit)},
  finally = {message("Finished model 2.")})
```

For the reliability diagrams, we compute the mean posterior predictive masses for each observation.
```{r}
#| label: classification probabilities
p_1 <- matrix(colMeans(fit_1$draws(variables = "ppm", format = "matrix")), ncol = K)
p_2 <- matrix(colMeans(fit_2$draws(variables = "ppm", format = "matrix")), ncol = K)
```

## PPC

The ppc bar plots don't show any major differences between the models.

:::{layout-ncol="2"}

```{r}
#| label: ppc_bars_1
#| fig-align: right
#| fig-width: 4
#| fig-height: 4
#| dev.args : 
#|   bg: transparent

ppc_bars(y = as.numeric(c),
         yrep = fit_1$draws(variables = "yrep",format = "matrix")) +
  ggtitle("Model 1") +
  theme(legend.position = "bottom")
```


```{r}
#| label: ppc_bars_2
#| fig-align: left
#| fig-width: 4
#| fig-height: 4
#| dev.args : 
#|   bg: transparent
ppc_bars(y = as.numeric(c),
         yrep = fit_2$draws(variables = "yrep", format = "matrix")) +
  ggtitle("Model 2") +
  theme(legend.position = "bottom") 
```
:::

When we inspect the reliability diagrams for the cumulative classification probabilities, we observe considerable under confidence with the first model. This is due to the omission to scale the mixture component variance when normalizing the data.
:::{layout-ncol="2"}
```{r}
#| label: reliability-diagrams_1
#| fig-align: right
#| fig-width: 4
#| fig-height: 4
#| dev.args : 
#|   bg: transparent

for (k in 1:(K - 1)) {
  p1 <- plot_dotted_reliabilitydiag(
    y = as.numeric(c <= k),
    x = if (k != 1) pmin(1, rowSums(p_1[, 1:k])) else p_1[, k],
    quantiles = K * N / 100,
    dot_scale = .5) +
    ggtitle(paste("Model 1: P(y <= ", k, ")", sep=""),
            subtitle = "1 dot = 100 observations")
  
  print(p1)
}
```

```{r}
#| label: reliability-diagrams_2
#| fig-align: left
#| fig-width: 4
#| fig-height: 4
#| dev.args : 
#|   bg: transparent

for (k in 1:(K - 1)) {
  p2 <- plot_dotted_reliabilitydiag(
    y = as.numeric(c <= k),
    x = if (k != 1) pmin(1, rowSums(p_2[, 1:k])) else p_2[, k],
    quantiles = K * N / 100,
    dot_scale = .5) +
    ggtitle(paste("Model 2: P(y <= ", k, ")", sep=""),
            subtitle = "1 dot = 100 observations")
  
  print(p2)
  }
```

:::





