---
title: "PPC Visualizations for Categorical Data"
subtitle: "Palmer Penguins"
date: "2023-01-30"
date-modified: "`r format(Sys.Date(), '%Y-%m-%d')`"
author: "Teemu Säilynoja"
image: "../../docs/case_studies/categorical/categorical_palmer_penguins_files/figure-html/groups_scatter-1.svg"
description: "Calibration plots for the easy tasks of identifying penguin species in the Palmer Penguins data set."
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    code-line-numbers: true
    default-image-extension: svg
    fig-format: svg
---



```{r inport, message=FALSE}

library("bayesplot")
library("cmdstanr")
library("ggplot2")
library("khroma")
library("quartoExtra")


# Source for the modified reliability plot
source("../../code/helpers.R")

theme_set(good_theme)
bayesplot_theme_set(good_theme)
color_scheme_set(scheme = c(unname(colour("vibrant")(7)[c(3,7,4,1,2,5)])))

scale_colour_discrete = scale_colour_vibrant
scale_fill_discrete = scale_fill_vibrant

source("../../code/helpers.R")

SAVE_MODEL = TRUE
```

Calibration plots for the easy tasks of identifying penguin species in the Palmer Penguins data set.

# The data

```{r load.data}
if (FALSE) {
  data("iris")
  X <- dplyr::select(na.omit(iris), -c("Species"))
  y <- as.numeric(iris$Species)
} else {
  library(palmerpenguins)
  data("penguins")
  X <- na.omit(penguins)[, c(3,4,5,6)]
  y <- as.factor(na.omit(penguins)$species)
}
```

```{r}
#| label: groups_scatter
#| dev.args :
#|   bg: transparent
ggplot(X, aes(x = bill_length_mm, y = bill_depth_mm, colour = y)) +
  geom_point() +
  xlab("Bill length (mm)") +
  ylab("Bill depth (mm)") +
  labs(colour = "Species") +
  legend_move(position = "top")
```

# The model

For the set of $N$ observations of $K$ features $x_{k,n}$ and their corresponding species $y_{n} \ in \{1,\dots,D\}$, we perform the classification using [the categorical logit classifier](https://mc-stan.org/docs/functions-reference/categorical-logit-glm.html), that is using the following likelihood
$$
p(y_n = s_j \mid \mathbf{w}) =  \frac{\exp\left(\beta_{n,j}\right)}{\sum_{d=1}^D\exp\left(\beta_{n,j}\right)},
$$
where
$$
\beta_{n,j} = w_{0,j} + \sum_{k=1}^Kw_{k,j}x_{k,n}
$$

::: {.callout-note appearance="simple" collapse="true" title="Stan code for the model"}
```{r}
model <- cmdstan_model(stan_file = "../../code/stan-models/penguins_glm.stan")
model
```
:::

After loading the model, we pass the data and run the inference using HMC.
As a result, we obtain 4000 samples from the posterior as well as posterior predictive samples and posterior predictive densities.

```{r}
fit <- model$sample(
  data = list(N = nrow(X),
              D = ncol(X),
              N_classes = length(unique(y)),
              X = X,
              y = as.numeric(y)),
  parallel_chains = 4,
  refresh = 0)
```

The posterior distributions of the weights show reasonable separation between classes (the first index). The dimension with most overlap seems to be the intercept term, $W[j,5]$ below.

```{r}
#| label: mcmc_areas_adelie
#| dev.args : 
#|   bg: transparent
mcmc_areas(fit$draws(variables = "W"))
```


# The calibration

The common approach of plotting a bar char of the observations overlaid with posterior means and 95% confidence intervals only gives a crude idea of the calibration of the model predictions.
```{r}
#| label: ppc_bars
#| message: false
#| dev.args : 
#|   bg: transparent
ppc_bars(as.numeric(y), fit$draws(variables = "yrep", format = "matrix")) +
  scale_x_continuous(breaks = 1:3, labels = levels(y))
```

From the following three reliability diagrams, one can see, that although sometimes underconfident in the predicted probabilities, the model is capable of very clearly separating the three species.

```{r}
#| label: reliab_adelie
#| dev.args : 
#|   bg: transparent
plot_dotted_reliabilitydiag(x = colMeans(exp(fit$draws(variables = paste(paste("lpd[", 1:nrow(X), sep=""), ",1]", sep=""), format = "matrix"))), y = as.numeric(y == levels(y)[1]), quantiles = 20) + labs(title = paste("Calibration:", levels(y)[1], "vs. Others"))

```


```{r}
#| label: reliab_chinstrap
#| dev.args : 
#|   bg: transparent
plot_dotted_reliabilitydiag(x = exp(colMeans(fit$draws(variables = paste(paste("lpd[", 1:nrow(X), sep=""), ",2]", sep=""), format = "matrix"))), y = as.numeric(y == levels(y)[2]), quantiles = 20) + labs(title = paste("Calibration:", levels(y)[2], "vs. Others"))
```


```{r}
#| label: reliab_gentoo
#| dev.args : 
#|   bg: transparent
plot_dotted_reliabilitydiag(x = exp(colMeans(fit$draws(variables = paste(paste("lpd[", 1:nrow(X), sep=""), ",3]", sep=""), format = "matrix"))), y = as.numeric(y == levels(y)[3]), quantiles = 20)  + labs(title = paste("Calibration:", levels(y)[3], "vs. Others"))
```



